\chapter{Introduction}
\label{introduction}

\section{Context and Motivation}
Wireless Sensor Networks (WSN) are ad-hoc networks that consist of sensor nodes that typically use low power radios such as IEEE 802.15.4, a relatively short range transmission standard radio technology in the 2.4 GHz band. The standard allows transmission to occur on several different channels within this band \cite{ieee802.15.4}. Unfortunately, the channels used by this technology often suffer interference \cite{interferenceModel, ieeeCompare}, for example, from Wi-Fi \cite{ieee_2012, wu} and Bluetooth. Sensor networks have to contend with an increasing number of devices that cause this wireless interference. Organising the network topology around this interference becomes an enabler for increasing transmission efficiency at a smaller energy cost. WSNs need to be able to operate reliably in the presence of such interference. It is important to minimise energy costs in these networks since deployments can be for weeks, months or longer.

Multichannel communication in wireless networks can alleviate the effects of interference which, as a result, can improve the network efficiency and stability, link reliability and minimise latency \cite{watteyne}. It also enables communication between physically proximate nodes to occur simultaneously without the risk of collision when the communicating nodes use different channels. However, not all channels are free from interference. Some channels would perform better than the other channels depending on the current location and network environment. Therefore, nodes should consider hopping to another channel when the channel shows an alarming decline in performance.
%thus, there is a gain to hop to another channel when the quality of the channel deteriorates. 
%Two commonly used types of channel hopping \cite{watteyne} are blind channel hopping and whitelisting. In blind channel hopping, nodes choose from all available channels. 
%Whitelisting, on the other hand, gives a set list of channels that avoids those that are known to commonly suffer interference.
%Many studies make use of channel whitelisting such as in Chrysso \cite{chrysso} and MiCMAC \cite{micmac}.

%Note that potentially Chrysso and MiCMAC could use all available channels.
%However, they do not have a mechanism to check the channel condition before using it for packet transmission. MiCMAC sees its performance degraded when using more than 4 channels, thus the decision on specifying 4 channels to be included in their experiment. 
%MiCMAC uses a different channel chosen at random each time it wakes up.
%It might require several wake up periods which is time consuming, before a clear channel is found from the 16 channels, to deliver the packet.  
%Chrysso on the other hand, switches the affected nodes to a new set of channels upon detecting interference which entails frequent channel switching if all channels are to be considered.

\section{Problem Statement}
It is clear that it is impossible to find a single channel guaranteed free from interference and there is no consensus on the best channel to use. The work in this PhD introduces a multichannel protocol that takes into account all available channels to utilise the spectrum and checks the condition of the channels before hopping to avoid those channels with interference. Several previous studies have developed multichannel MAC layer but, despite the potential benefits none are yet widely implemented in real world deployments.

\section{Contribution}
Important aspects of this work is in improving the WSNs performance by implementing multichannel routing protocol to investigate the effect in comparison to a single channel in a noisy and lossy environment. There are several research challenges in a multichannel protocol design in term of the synchronisation, topology formation and maintenance, channels initiation, channel selection processes and the energy consumption of the multichannel protocol. The multichannel protocol in this work consists of centralised and decentralised parts where the centralised node controls the network and the decentralised node carries out the multi channel processes. The work in this PhD investigates the channel selection processes based on the nodes cross layers interactions during decision making.

%The decision making process is (centralized and decentralized – explain here). The main benefits of this approach are (). 
%The work in this PhD will address the issues raised by (). More specifically it will investigate how the channel selection is determined from the nodes interactions. In addition, it will investigate the cross layer interaction. 

\section{Current Work}
Multichannel Cross-Layer Routing Protocol (MCRP) has been developed as a new multichannel protocol. MCRP is proposed to enable communications on all available channels in the spectrum to avoid interference, congestion and conflict in the network. MCRP consists of two main parts; a centralised intelligence at LPBR, and decentralised nodes. LPBR implements a two-hop colouring algorithm to avoid interference between physically proximate nodes trying to communicate on the same channel. The information on channel interference and network topology from the lower layer is made available to the application layer. This allows the centralised controller (LPBR) to have an overall view of the system to make decisions at the network and MAC layers about which channels nodes should listen on. The system is fail safe in the sense that the WSN functions if the central system which assigns channels fails temporarily or permanently. MCRP avoids channels with interference which greatly reduces the effects of interference on the network. MCRP is implemented in Contiki \cite{contiki}, an open source operating system for WSNs. The performance of MCRP has been evaluated in Contiki network simulator, Cooja \cite{cooja} and Flocklab \cite{flocklab} testbed.

%A new multichannel protocol called Multichannel Cross-Layer Routing Protocol (MCRP) has been developed. The proposed approach in MCRP is that the nodes are able to communicate on many channels in order to avoid interference and channel congestion in a centralised and decentralised manner.
%This paper presents a Multichannel Cross-Layer Routing Protocol (MCRP) which 
%We demonstrate that 
%%The performance of the approach has been evaluated using (). The evaluation has been performed with respect to the ().
%%In the context of the research efforts carried out from the beginning of the PhD research work, the followings have been investigated.


\section{Report Outline}
The remainder of the report is organised as follows. Chapter 2 introduces the state-of-the art in the area of multichannel protocols. It also presents the main current research efforts towards () Section \ref{sec:relatedwork} presents related work to multichannel protocols. Chapter 3 presents the main features and mechanisms used in MCRP. It describes (). It also presents (). Section \ref{sec:multichannel} describes the key idea of our proposed protocol and the high-level design, and the implementation of the protocol in Contiki. We describe and evaluate the experimental results in Section \ref{sec:evaluation}. Chapter 7 summarises the current work and presents the future research works that will be investigated in the context of this PhD.